---
title: "Time series data wrangling, exploration and visualization"
author: "Madeline Berger"
date: "2020-03-10"
output: html_document
summary: 'Examining steelhead salmon crossings at the Bonneville Dam'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)

# For general stuff:
library(tidyverse)
library(janitor)
library(lubridate)
library(here)
library(paletteer)

# For ts stuff: 
library(tsibble)
library(fable)
library(fabletools)
library(feasts)


steelhead <- read_csv("steelhead_tidy.csv") %>% 
  clean_names()
```



***
##Introduction 


This project uses data on steelhead salmon fish passage at the Bonneville Dam on the Columbia River to explore time series data cleaning and visualization. The data was compiled by the U.S. Army Corps of Engineers and includes a value for daily number of fish that passed through the dam from 1940 to 2019. Also included is the data with a column for year and another column for month and day. 


*References*  
1. Columbia River DART, Columbia Basin Research, University of Washington. (2019). Adult Passage Graphics & Text. Available from http://www.cbr.washington.edu/dart/query/adult_graph_text  

2. Photo by Dave Rotb on Unsplash
  
***  

##Analysis

###Data Cleaning and Wrangling

To better visualize this data in multiple ways, the dates first need to be parsed into date format and cleaned. They are read in as character data, which will not work logically when graphed. 

```{r}

#check out some of the columns - are they all necessary?

#unique(steelhead$datatype) #only one data type

#min(steelhead$year)

#first want to separate out the dates and parse data
steelhead_clean <- steelhead %>%
  mutate(
    date = paste(year, mm_dd, sep = "-")
    ) %>% 
  mutate(
    yr_mo_day = lubridate::parse_date_time(date, "ydm"),
    yr_mo_day = as.Date(yr_mo_day), #since ours started out as characters
    month_sep = yearmonth(yr_mo_day)
  ) %>% 
  mutate(
    month = month(yr_mo_day, label = TRUE),
    value = as.numeric(value)
  ) %>% 
  select(-mm_dd, -datatype, -location, -parameter) %>% 
  drop_na(month_sep, value)
#check_NAs

#steelhead_dates_final <- steelhead_dates[,c(4,5,1,3,2)]


```


###Daily Steelhead Passage
Visualize daily passage with a time series plot:
```{r}
#daily data is challenging - plot the whole yr_mo_day

steelhead_daily <- ggplot(steelhead_clean, aes(x = yr_mo_day, y = value))+
  geom_line(color = "royalblue1") + 
  theme_minimal()+
  labs(x = "Date", y = "Steelhead salmon passge")

steelhead_daily

```
  
    

**Figure 1. Daily steelhead salmon counts passing through Bonneville Dam on the Columbia River.** Counts were recorded daily from 1940 to 2019.


###Monthly steelhead passage
This can be visualized with a seasonplot:
```{r}
#group by month
steelhead_by_month <- steelhead_clean %>%
  group_by(month, year) %>% 
  summarize(
    month_tot = sum(value)
  ) 


#plot 
steelhead_seasonplot <- ggplot(data = steelhead_by_month , aes(x = month, y = month_tot, group = year)) +
   geom_line(aes(color = year))+
  theme_minimal()+
  scale_y_continuous(expand = c(0,0), breaks = seq(0, 400000, by = 50000), labels = scales::comma)+
  labs(x = "Month", y = "Total steelhead salmon passage", color = "Year")


steelhead_seasonplot +
  scale_color_paletteer_c("pals::warmcool")

```
  
    
**Figure 2. Monthly steelhead salmon passge at the Bonneville Dam on the Columbia River.** Each line depicts a different year, with red colors indicating earlier recordings and blue colors indicating more recent data. 


###Annual Steelhead passage

```{r}
#group by year
steelhead_by_year <- steelhead_clean %>%
  group_by(year) %>% 
  summarize(
    yearly_tot = sum(value)
  ) 


steelhead_yearly <- ggplot(steelhead_by_year, aes(x = year, y = yearly_tot))+
  geom_line(color = "salmon")+
  scale_y_continuous(expand = c(0,0), breaks = seq(0, 600000, by = 100000), labels = scales::comma)+
  theme_minimal()+
  labs(x = "Year", y = "Steelhead salmon passage")

steelhead_yearly

```
  
    

**Figure 3. Yearly steelhead salmon trends at Bonneville Dam, Orgeon.** The y-axis depicts the total number  of fish counted at the dam for each year, ranging from 1940 to 2019. 

***

##Results

From these visualizations, we can see how steelhead numbers in the Columbia River have varied each year and each season. The daily passage time series plot indicates that daily steelhead counts have been somewhat consistent over time following a cyclical pattern. The exception is one day in 2009 that had very high counts.

The season plot helps illuminate what may be driving some of the cyclical trends in the time series plot. Every year, there is a clear increase in steelhead salmon passage starting from June to October, peaking dramatically in August. However, we can see there have been differences in the magnitude of this peak over the years. There appears to be an increasing trend from 1940 - 2019.

The annual passage plot confirms this observation, showing that total yearly counts of steelhead started increasing in 1980, but then have decreased since 2010. 

